\section{Type Checker Generators}
There have been multiple attempts to make the writing of type checkers easier on the compiler writers by creating new tools or notations.
These can have very different ways of  expressing type systems and what kinds of notations they use.
Some use the notation of attribute grammars to represent type systems, along with the rest of the compiler, whereas other use more specialised DSL's to represent their type rules and type checkers.

\subsection{Attribute Grammars}
Attribute Grammars are a way of specifying the semantics of grammar rules for a Context Free Grammar.
This makes them a likely candidate to look at when attempting to generate type checking code since type checking is a form of semantic analysis on some input code.
In Knuth's paper\cite{Knuth1968}, he gives an example of the attributes required to evaluate a simple language to emulate a Turing Machine.
Even though he discusses the evaluation of a turing machine, they are not limited to expressing the semantics of evaluation, we can also use them to express the semantics of type systems such that on each node of the tree an attribute could be given to it that defines the type of the node.

\subsection{Eli}
\begin{figure}[]
    \centering
    \begin{prooftree}
        \AxiomC{$\Gamma \vdash e_1 : \tau$}
        \AxiomC{$\Gamma \vdash e_2 : \tau$}
        \RightLabel{$[\tau \in \{Bool, Int\}]$}
        \BinaryInfC{$\Gamma \vdash e_1 = e_2 : Bool$}
    \end{prooftree}
    \caption{Type rule for equality for the example language for Eli}
    \label{fig:oilNatRule}
\end{figure}
Eli is a tool which, given a set of specifications, will generate code for a whole compiler based upon those specifications\cite{Gray:1992:ECF:129630.129637}.
The way Eli generates all the code is by the use of an attribute grammar system and numerous internal tools that generate code for each different part of the compiler pipeline.
The internal tools use DSL's which define a specification for part of the language.
One of these DSL's is called \textit{Operator Identification Language} (OIL), which is used to define the typing relations on operators.
The definition of these type rules is fairly simple, however in their example there are a lot of repeated definitions for operators.
For example there are multiple definitions to define the equality operator, one for each of the required types that the operator can act on.
Where as in the natural deduction rules this would only have to be defined once where the type is some arbitrary type $\tau$  with some constraint on $\tau$ such that it can only be from a set of types that an equality check can be performed on, see the natural deduction rule given in \autoref{fig:oilNatRule}.
The tools lacks built-in support for constructs in the language that can have arbitrary types.
However, OIL is a small and simple language that is easy to understand, but, the other constituent parts of ELI required to build a tool are not as simple and the specifications written in them can get significantly more complicated for larger projects.

\subsection{Ruler}
Ruler is a DSL that bears close resemblance to natural deduction rules that formally define type systems\cite{dijkstra2006ruler}.
The notation used to describe the type rules in ruler is intuitive to anyone who has experience with using natural deduction rules to represent type rules.
However, other possible compiler writers may have never come across the notation therefore the syntax of Ruler could be un-intuitive to them.
The rules can be compiled to either \LaTeX or to an attribute grammar system in Haskell.
However, this means that rules have to be written multiple times for each target that the user may want to output to.
This creates unnecessary duplication of types rules since if a user wants both the executable code and the formal specification they will have to write the type system twice for both of these outputs.
Since the only executable code that is generated compiles to an attribute grammar, it forces the user to use the attribute grammar for the whole frontend of their compiler, whereas using the abstract syntax allows the user to choose any parser generator (or write their own parser) as long they provide the abstract syntax to the tool.
They also make a claim that they support Hindley-Milner type inference.
While this to some extent is true, it requires the user to write a rather large amount of Haskell code in order for the tool to generate correct code for this type inference.
This means that a large part of the type checker still has to be written by hand because many non-trivial type rules will require access to the context or for types to be unified (this is a required process for Hindley-Milner type inference) for which code is not automatically generated.

\subsection{Typical}
\label{sec:typical}
The final tool we will be discussing is Typical\cite{grimm2007typical}.
Typical is a language that uses ML\cite{milner1997definition} as the base of its language.
This means that all terms in Typical are typed and such as a proof of concept of Typical, the type checker for Typical is built using Typical.
Typical also has some extra declarations that are not in ML, these are the ability to define name-spaces for the context and scoping rules.
In the tools we have discusses so far the users have had to write such features of their type system.
Typical also has in built functions that retrieve types from the context and adds new items to a context, which is again features that the previous tools did not have.
Unlike our previous examples that use attribute grammars to perform their type checking, Typical uses the abstract syntax tree given by the parser.
This means that Typical can be used with any parser generator that targets Typical's target language (Java), although it is stated that some modification may be required.
The required modification has been performed on the Rats! parser generator\cite{Grimm:2006:BET:1133255.1133987}.
To understand a type system written in Typical one must be familiar with ML or another language similar to it, such as Haskell, since the vast majority of the code in Typical is ML.